---
title: "ggplot2 demystified - iDiv tutorial"
author: "Petr Keil"
date: "May 2, 2018"
output: 
  html_document: 
    fig_height: 4
    fig_width: 6
    highlight: pygments
    number_sections: yes
    theme: cerulean
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For a pretty version click [here](https://rawgit.com/petrkeil/ggplot2_demystified/master/ggplot2_demystified.html).

# Introduction

This is a tutorial to basic `ggplot2` syntax and functionality. It will be used
during a 1h class for PhD students and postdocs at iDiv.

The main goals are:

- To explain the basic (and unusual) `ggplot2` syntax, and to clarify terms such as 
`geom`, `aes`, `theme`, `stats`, `scale`.

- To give examples of the powerful `ggplot2` functionality, with special 
emphasis on **multivariate data**.

## Why `ggplot2`? What is it?

- An R package for data visualization based on [The Grammar of Graphics](https://www.amazon.com/Grammar-Graphics-Statistics-Computing/dp/0387245448/ref=as_li_ss_tl?ie=UTF8&qid=1477928463&sr=8-1&keywords=the+grammar+of+graphics&linkCode=sl1&tag=ggplot2-20&linkId=f0130e557161b83fbe97ba0e9175c431)

- Created by Hadley Wickham

- Inspired by good things from both `graphics` and `lattice`

- Strong underlying model -- makes things concise, consistent, readable, and flexible

- Great especially for mutlidimensional data (simulations, large observational data)

- Excellent documentation, tutorials, widespread use

- Beware: steep learning curve

## Documentation

The best one is [here](http://ggplot2.tidyverse.org/reference/).

## Load the library

```{r}
library(ggplot2)
```

## Example data

We will use global data on **tiger beetles** from Cassola and Pearson (2000) *Biological Conservation* 95: 197-208.

```{r}
beetles <- read.csv("http://www.petrkeil.com/wp-content/uploads/2017/09/tiger_beetles.csv") 
summary(beetles)
```

# Basic syntax and elements

Most important things: `geom`, `aes`, `scale`, `stat`, `facet`

## `qplot`

`qplot` is simple to use, designed to produce easy plots, but doesn't have the full capability.

```{r}
qplot(x = Area, y = S, data=beetles)
```

## `ggplot`

In the rest of the intro we will focus on `ggplot`. It's always the same syntax:

1. Initialize `ggplot`, define data and the x and y

2. Once ggplot is intialized, add stuff (geoms, stats, scales, themes):


```{r}
p <- ggplot(data = beetles, aes(x=Area, y=S)) 
p
```

## `geom`

Describes type of geometric objects to plot (points, lines, bars, ...)

```{r}
p <- ggplot(data = beetles, aes(x=Area, y=S)) +
     geom_point()
p
```

## `scale`

Describes how the data are converted to display values (log scales, color scales, ...)

```{r}
p <- ggplot(data = beetles, aes(x=Area, y=S)) + 
     geom_point() +
     scale_x_log10() + scale_y_log10()
p
```

## `aes`

Describes properties of geoms that **vary** (position, colour, size, transparency)


```{r}
p <- ggplot(data = beetles, aes(x=Area, y=S)) + 
     geom_point(aes(colour = Endemics)) + 
     scale_x_log10() + scale_y_log10()
p
```

```{r}
p <- ggplot(data = beetles, aes(x=Area, y=S)) + 
     geom_point(aes(size = Endemics, colour = Endemics)) + 
     scale_x_log10() + scale_y_log10()
p
```

```{r}
p <- ggplot(data = beetles, aes(x=Area, y=S)) + 
     geom_point(aes(colour = Realm)) + 
     scale_x_log10() + scale_y_log10()
p
```

## `stat`

Does statistical transformations or models (means, medians, regression lines)

```{r}
p <- ggplot(data = beetles, aes(x=Area, y=S)) + 
     geom_point(aes(colour = Realm)) + 
     scale_x_log10() + scale_y_log10() +
     stat_smooth(aes(colour = Realm), method="lm", se=FALSE) 
p
```

## `facet`

```{r, fig.width = 9, fig.height = 8}
p <- ggplot(data = beetles, aes(x=Area, y=S)) + 
     geom_point(aes(colour = Realm)) + 
     stat_smooth(aes(colour = Realm), method="lm", se=FALSE) +
     scale_x_log10() + scale_y_log10() + 
     facet_wrap( ~ Realm)
p
```

```{r, fig.width = 9, fig.height = 8}
p <- ggplot(data = beetles, aes(x=Area, y=S)) + 
     geom_point(aes(colour = Realm, size = Endemics)) + 
     stat_smooth(aes(colour = Realm), method="lm", se=FALSE) +
     scale_x_log10() + scale_y_log10() + 
     facet_wrap( ~ Realm)
p
```

```{r, fig.width = 9, fig.height = 8}
p <- ggplot(data = beetles, aes(x=Area, y=S)) + 
     # To repeat the same data in every panel, simply construct a data frame
     # that does not contain the facetting variable.
     geom_point(data = transform(beetles, Realm = NULL), colour = "grey85", aes(size= Endemics)) +
     geom_point(aes(colour = Realm, size = Endemics)) + 
     stat_smooth(aes(colour = Realm), method="lm", se=FALSE) +
     scale_x_log10() + scale_y_log10() + 
     facet_wrap( ~ Realm)
p
```

## `theme`

```{r, fig.width = 9, fig.height = 8}
p <- ggplot(data = beetles, aes(x=Area, y=S)) + 
     geom_point(data = transform(beetles, Realm = NULL), colour = "grey85", aes(size= Endemics)) +
     geom_point(aes(colour = Realm, size = Endemics)) + 
     stat_smooth(aes(colour = Realm), method="lm", se=FALSE) +
     scale_x_log10() + scale_y_log10() + 
     facet_wrap( ~ Realm) + 
     theme_bw() # try theme_classic, theme_minimal, or package ggthemes
p
```

# More tricks

## Boxplots

```{r}
p.box <- ggplot(data = beetles, aes(x=Realm, y=S)) + 
         geom_boxplot(aes(fill = Realm)) + 
         scale_y_log10() 
p.box
```

## Histograms

```{r}
p.hist <- ggplot(data = beetles, aes(x=S)) + 
          geom_histogram() +
          scale_x_log10()
p.hist
```

## Multiple panels

```{r, fig.width=10}
library(gridExtra)
grid.arrange(p.hist, p.box, ncol=2)
```

## Flipping coordinate systems

```{r, fig.width=10}
grid.arrange(p.hist , p.box + coord_flip(), ncol=2)
```

## Colour brewer

```{r}
p.box <- ggplot(data = beetles, aes(x=Realm, y=S)) + 
         geom_boxplot(aes(fill = Realm)) + 
         scale_y_log10() +
         scale_fill_brewer(palette="Set2")
p.box
```

## Annotation

```{r}
p.box <- ggplot(data = beetles, aes(x=Realm, y=S)) + 
         geom_boxplot(aes(fill = Realm)) + 
         scale_y_log10() +
         scale_fill_brewer(palette="Set2") +
         labs(title="A", 
              subtitle = "Colourful boxplot",
              y="Species richness", 
              caption = "Copyright to Bored Badger.")
p.box
```




